---
title: "DATA607 FINAL PROJECT"
subtitle: "NCAA Athletics and College Finances"
author: "Logan Thomson"
date: "May 5, 2016"
output:
  slidy_presentation: 
    fig_width: 14
    fig_height: 9
---

# Introduction  

```{r load_packages, echo=FALSE, message=FALSE, results='hide'}
packages <- c("tidyr", "dplyr", "ggplot2", "XML", "xml2", "ggmap")
lapply(packages, library, character.only = T)
```

“A very small number of the 1,100 (NCAA members) have a positive cash flow on college sports, so those schools are making a decision that having a successful athletic program is valuable to them despite the fact they have to subsidize it with institutional money. The same thing is true for a lot of academic programs. So every school has to sit down and say, 'What is this worth to us?’ ” - NCAA President & CEO Mark Emmert  

- Are college athletics a huge source of revenue?
- Do college athletics pose a financial burden?
- Is there a positive or negative effect on institutions?

# Data Sources

Data                    | Type | Variables                                | Source  
----------------------- | ---- | ---------------------------------------- | ------  
NCAA APR | .tsv | Conference, Sub-Divisions, Sports, Academic Performance | http://www.icpsr.umich.edu/icpsrweb/NCAA/studies/26801  
Knight Foundation | .csv | Academic & Athletic Spending | http://spendingdatabase.knightcommission.org  
USA Today NCAA Finances | Scraped Web Page | Athletic Expenses & Revenues | http://sports.usatoday.com/ncaa/finances  
IPEDS | .csv | Admission, Retention, Graduation | http://nces.ed.gov/ipeds/datacenter/Default.aspx  

```{r icpsr, echo=FALSE}
# open file
path <- ("https://raw.githubusercontent.com/Logan213/DATA607_Final/master/NCAA_ICPSR.tsv")
con <- file(path, open="r")

# NCAA ICPSR data for conferences, sub-divisions, and number of sports
icpsr <- read.csv(con, header=T, sep="\t", stringsAsFactors = F)

# close file
close(con)

# Select only 2014 data
icpsr <- icpsr %>%
  select(SCL_UNITID, SCL_NAME, DIV=SCL_DIV_14, SUB_DIV=SCL_SUB_14, CONFERENCE=CONFNAME_14, PUB_PRIV=SCL_PRIVATE) %>%
  filter(SUB_DIV > 0) %>%  # Filter out -99 values
  group_by(SCL_UNITID, SCL_NAME, DIV, SUB_DIV, CONFERENCE, PUB_PRIV) %>%
  summarise(NUM_SPORTS=n())  # Count number of each sport
```  

```{r knight_found, echo=FALSE}

# Knight Foundation Data

# open file
path <- ("https://raw.githubusercontent.com/Logan213/DATA607_Final/master/knight.found.csv")
con <- file(path, open="r")

knight <- read.csv(con, header=T, stringsAsFactors = F)

# close file
close(con)

knight <- knight[c(1:5, 7, 9:10, 12, 14, 18, 20, 22, 24, 26, 32, 35:37)]
knight <- knight[knight$Year == 2014, ]
knight <- knight[1:231,]
```  

```{r ipeds, echo=FALSE}

# IPEDS Data

# open file
path <- ("https://raw.githubusercontent.com/Logan213/DATA607_Final/master/IPEDS_2014.csv")
con <- file(path, open="r")

ipeds <- read.csv('~/desktop/IPEDS_2014.csv', header=T, stringsAsFactors = F)

# close file
close(con)

# Select and rename columns
ipeds <- ipeds[c(1:6, 8:10)]

colnames(ipeds) <- c("unitid", "Inst_Name", "Year", "FT_Retention", "Grad_Rate", "Admission_Rate", "City", "State", "Zip_Code")

#remove "+4" postal codes
ipeds$Zip_Code <- substring(ipeds$Zip_Code, 1, 5)
```


```{r usa_today, echo=FALSE}
# USA Today NCAA Finance "database" for expenses and revenues

html_url <- "http://sports.usatoday.com/ncaa/finances"
html_url <- read_html(html_url)  #Store url

parsed_ncaa <- htmlParse(html_url)  # Parse url and use readHTMLTable
ncaa_exp_rev <- readHTMLTable(parsed_ncaa)

ncaa_exp_rev <- data.frame(ncaa_exp_rev$`sports-table-scroll-dom`, stringsAsFactors = FALSE)  # Convert to a data frame

# Convert to appropriate columns to character and number
ncaa_exp_rev[3:6] <- lapply(ncaa_exp_rev[3:6], function(x) as.character(x))
ncaa_exp_rev[3:6] <- lapply(ncaa_exp_rev[3:6], function(x) as.numeric(gsub('[^0-9.]', '', x)))
```  

```{r ipeds_codes, echo=FALSE}

# Could not get regex to work for merging on names

# Manually entered column of IPEDS ids to merge data frames on
ncaa_exp_rev$ipeds <- as.numeric(c(228723, 228778, 204796, 170976, 100751, 134130, 159391, 207500, 221759, 214777, 100858, 240444, 134097, 157085, 139959, 106397, 218663, 174066, 171100, 153658, 209551, 157289, 236948, 181464, 110662, 207388, 163286, 155317, 234076, 178396, 238032, 199120, 151351, 176017, 104179, 145637, 110635, 104151, 217882, 233921, 229115, 139755, 199193, 243780, 176080, 155399, 153603, 129020, 186380, 126614, 209542, 230764, 236939, 201885, 132903, 128328, 122409, 198464, 137351, 182281, 232423, 225511, 232982, 142115, 220862, 141574, 187985, " ", 110556, 126818, 166629, 240727, 172699, 228459, 169798, 110644, 199139, 204024, 130943, 196088, 100663, 200800, 220978, 234030, 227216, 133669, 230728, 157951, 228796, 197869, 106458, 196097, 139940, 204857, 133951, 122755, 102094, 169248, 183044, 147703, 237525, 102368, 188030, 182290, 229027, 156125, 206084, 145813, 110422, 232186, 203517, 218724, 217484, 231624, 179566, 164076, 200280, 176372, 160658, 150136, 180489, 149222, 200332, 159647, 110617, 201441, 139931, 140164, 154095, 161253, 142285, 217819, 180461, 231174, 196060, 110653, 105330, 110583, 221740, 110705, 228431, 219356, 181394, 145600, 110671, 232937, 199148, 226091, 166513, 220075, 227881, 127741, 157401, 163268, 196079, 217864 ,128771, 206695, 219471, 433660, 221847, 100724, 151324, 156620, 185828, 178402, 110608, 171571, 199218, 110565, 240453, 235097, 144892, 110486, 157447, 101480, 209807, 149772, 159993, 160621, 234085, 221838, 233277, 228769, 130934, 106704, 218964, 133650, 160612, 142276, 200004, 230737, 230782, 199102, 202134, 199157, 179557, 157386, 136172, 224147, 219602, 159717, 230603, 160038, 206604, 221768, 227368, 227526, 163453, 100654, 229063, 232566, 106245, 151111, 218733, 149231, 240277, 175856, 159966, 151102, 218742, 175342, 199111, 106412, 144005, 140960, 163338, 159009, 159939, 176044, 162283))
```

# Overview of Combined Data

```{r merge_df, echo=FALSE}
# Merge Data Frames
ncaa <- merge(x=ncaa_exp_rev, y=ipeds, by.x="ipeds", by.y="unitid")
ncaa <- merge(x=ncaa, y=knight, by.x="ipeds", by.y="IPEDS.ID")
ncaa <- merge(x=ncaa, y=icpsr, by.x="ipeds", by.y="SCL_UNITID")

# Still a lot of columns, paring some down and renaming
ncaa <- ncaa %>%
  select(IPEDS_ID=ipeds, 
         TOT_ATH_REVENUE=total.revenue, TOT_ATH_EXPENSE=total.expenses,
         ATH_SUBSIDY=total.subsidy, PCT_SUBSIDY=X..subsidy,
         FT_RET_RATE=FT_Retention, Grad_Rate, Admit_Rate=Admission_Rate,
         City, State, Zip_Code, 
         Institution=Data, FTE.Students, 
         TOT_ACAD_SPEND=Total.Academic.Spending,
         TOT_INSTRUC_SPEND=Total.Instructional.Spending,
         NUM_ATHLETES=Unduplicated.Athletes, 
         TOT_ATH_SPEND=Total.Athletic.Spending,
         TOT_FB_SPEND=Total.Football.Spending,
         TOT_ATH_FUNDING=Total.Institutional.Funding.for.Athletics,
         COACH_SAL=Total.Coaching.Salaries, 
         FB_PLYRS=Football.Players, 
         FB_COACH_SAL=Total.Football.Coaching.Salaries,
         YR_DEBT_SERV=Annual.Debt.Service.on.Athletic.Facilities,
         TOT_ATH_DEBT=Total.Debt.Outstanding.on.Athletic.Facilities, 
         DIV, SUB_DIV, CONFERENCE, PUB_PRIV, NUM_SPORTS)

# Reorder columns
ncaa <- ncaa[c(1,12,9,10,11,28,13,16,21,8,6,7,25:27,29,14:15,17:18,20,22,19,23:24,2:5)]

# Remove dollar signs, commas, and convert currencies to numeric
ncaa[17:25] <- lapply(ncaa[17:25], function(x) as.numeric(gsub('[^0-9]', '', x)))
```  

- Division I institutions only

- Limited to financial data in Knight Foundation and USA Today 

- 29 different variables

- Data on 225 institutions

----

```{r ggmap, echo=FALSE, message=FALSE, cache=TRUE, warning=FALSE}
# plot athletic spending on map
ncaa_mapping <- ncaa %>%
  select(Institution, Zip_Code, CONFERENCE, TOT_ATH_SPEND) %>%
  filter(Zip_Code != "96822")

for (i in 1:nrow(ncaa_mapping)) {  
  latlon = geocode(ncaa_mapping[i,2])
  ncaa_mapping$lon[i] = as.numeric(latlon[1])
  ncaa_mapping$lat[i] = as.numeric(latlon[2])
}

us_center = as.numeric(geocode("United States"))

us_map = ggmap(get_googlemap(center=us_center, color="bw", scale=2, zoom=4), extent="panel")

us_map +  
   geom_point(aes(x=lon, y=lat), data=ncaa_mapping, col="deepskyblue4", alpha=0.3, size=ncaa_mapping$TOT_ATH_SPEND/8500000) +  
   scale_size_continuous(range=range(ncaa_mapping$TOT_ATH_SPEND)) + 
  labs(x="Longitude", y="Latitude") + ggtitle("NCAA Div. I Institutions: Athletic Spending")
```  

```{r millions, echo=FALSE}
# For re-sizing graph axes
mil <- 1000000
```

----

```{r num_teams, echo=FALSE}

# Number of Sports Teams

ncaa %>%
  ggplot(aes(NUM_SPORTS, fill=factor(SUB_DIV))) + 
  geom_histogram(bins=20, alpha=0.6) +
  labs(title="NCAA Sports Per Institution", fill="Div. I Subdivision", x="Number of Sports", y="Count") +
  scale_fill_discrete(breaks=c(1,2,3), labels=c("FBS", "FCS", "w/o FB")) + 
    theme(axis.text.x=element_text(size=12), axis.title.x=element_text(size=14),
        axis.text.y=element_text(size=12), axis.title.y=element_text(size=14),
        plot.title=element_text(size=20, face="bold"), legend.position = "bottom")
```  

----

```{r ath_per_student, echo=FALSE}

# Number of Athletes per Student

ncaa %>%
  ggplot(aes(x=FTE.Students, y=NUM_ATHLETES)) +
  geom_point(aes(size=NUM_SPORTS, color=factor(SUB_DIV)), alpha=.65) +
  labs(title="Athletes per FT Enrolled Student", color="Div. I Subdivision", x="FTE Students", y="Unduplicated Athletes") +
  scale_color_discrete(breaks=c(1,2,3), labels=c("FBS", "FCS", "w/o FB")) +
  theme(axis.text.x=element_text(size=12), axis.title.x=element_text(size=14),
        axis.text.y=element_text(size=12), axis.title.y=element_text(size=14),
        plot.title=element_text(size=20, face="bold"), legend.position = "bottom")
```  

----

```{r acad_v_ath, echo=FALSE, warning=FALSE}

# Academic Spend vs. Ath Spend

ncaa %>% 
  ggplot(aes(x=(TOT_ACAD_SPEND/mil), y=(TOT_ATH_SPEND/mil), color=factor(SUB_DIV)), na.rm=T) + 
  geom_point(aes(size=FTE.Students), alpha=0.6) + 
  geom_smooth(method="lm", se=F, color='black') +
  scale_size(breaks=c(10000, 20000, 30000, 40000, 50000)) +
  scale_y_continuous(labels = scales::dollar) + 
  #scale_x_continuous(labels = scales::dollar) + 
  labs(title="Academic Spending vs. Athletic Spending", color="Div. I Subdivision", x="Total Academic Spending (in Millions)", y="Total Athletic Spending (in Millions)") + 
  scale_color_discrete(breaks=c(1,2,3), labels=c("FBS", "FCS", "w/o FB")) +
  theme(axis.text.x=element_text(size=12), axis.title.x=element_text(size=14),
        axis.text.y=element_text(size=12), axis.title.y=element_text(size=14),
        plot.title=element_text(size=20, face="bold"), legend.position = "bottom")
```  

----

```{r enroll_v_ath, echo=FALSE, warning=FALSE}

# Enrollment vs. Athletic Spending

ncaa %>%
  ggplot(aes(x=FTE.Students, y=TOT_ATH_SPEND/mil)) + 
  geom_point(aes(color=factor(SUB_DIV)), size=4, alpha=0.6) + 
  labs(title="FT Enrollment vs. Athletic Spending", color="Div. I Subdivision", x="FTE Students", y="TOTAL ATH. SPENDING (in Millions)") + 
  geom_smooth(method="lm", se=F, color='black') + 
  scale_y_continuous(labels = scales::dollar) + 
  theme(axis.text.x=element_text(size=12), axis.title.x=element_text(size=14),
        axis.text.y=element_text(size=12), axis.title.y=element_text(size=14),
        plot.title=element_text(size=20, face="bold"), legend.position = "bottom")
``` 


----

```{r pwr_five, echo=FALSE}

# Expenses vs. Revenue

ncaa %>% 
  ggplot(aes(x=TOT_ATH_REVENUE/mil, y=TOT_ATH_EXPENSE/mil)) +
  geom_point(aes(size=FTE.Students, color=factor(SUB_DIV)), alpha=.6) + 
  scale_y_continuous(labels = scales::dollar) +
  scale_x_continuous(labels = scales::dollar) + 
  labs(title="Atheletic Expense vs. Revenue", color="Div. I Subdivision", x="Total Athletic Revenue (in Millions)", y="Total Athletic Expense (in Millions)") + 
  theme(axis.text.x=element_text(size=12), axis.title.x=element_text(size=14),
        axis.text.y=element_text(size=12), axis.title.y=element_text(size=14),
        plot.title=element_text(size=20, face="bold"), legend.position = "bottom")
```  

----

```{r oth_conf, echo=FALSE}

# No FBS Schools, Revenue vs. Expenses

no_fbs <- ncaa %>% 
  filter(SUB_DIV != 1)

no_fbs %>%
  ggplot(aes(x=TOT_ATH_REVENUE/mil, y=TOT_ATH_EXPENSE/mil)) +
  geom_point(aes(size=FTE.Students, color=factor(SUB_DIV)), alpha=.6) + 
  scale_y_continuous(labels = scales::dollar) +
  scale_x_continuous(labels = scales::dollar) + 
  labs(title="Atheletic Expense vs. Revenue: No FBS", color="Div. I Subdivision", x="Total Athletic Revenue (in Millions)", y="Total Athletic Expense (in Millions)") + 
  theme(axis.text.x=element_text(size=12), axis.title.x=element_text(size=14),
        axis.text.y=element_text(size=12), axis.title.y=element_text(size=14),
        plot.title=element_text(size=20, face="bold"), legend.position = "bottom")
```  

----

```{r all_adm_pct, echo=FALSE, warning=FALSE}

# Difference in Athletic Rev/Exp vs. Selectivity, All Inst.

ncaa %>%
  ggplot(aes(x=Admit_Rate, y=((TOT_ATH_REVENUE/mil)-(TOT_ATH_EXPENSE/mil)), fill=CONFERENCE)) +
  geom_bar(stat='identity', position='identity', color='black', alpha=0.7) +
  labs(title="Athletic Expense vs. Admission Selectivity", x="Admission Accept %", y="Athletic Rev - Expense (in Millions)") +
  scale_y_continuous(labels = scales::dollar) +
  theme(axis.text.x=element_text(size=12), axis.title.x=element_text(size=14),
        axis.text.y=element_text(size=12), axis.title.y=element_text(size=14),
        plot.title=element_text(size=20, face="bold"), legend.position = "bottom")
```  

----

```{r conf_split, echo=FALSE}

# Split up "Power 5" and everyone else

pwr_five <- ncaa %>%
  filter(grepl("Atlantic Coast Conference|Big 12 Conference|Big Ten Conference|Southeastern Conference|Pac-12 Conference", CONFERENCE))

other_conf <- ncaa %>%
  filter(!grepl("Atlantic Coast Conference|Big 12 Conference|Big Ten Conference|Southeastern Conference|Pac-12 Conference", CONFERENCE))
```

```{r pwr_five_adm_pct, echo=FALSE}

# Difference in Athletic Rev/Exp vs. Selectivity, "Power 5"

pwr_five %>%
  ggplot(aes(x=Admit_Rate, y=((TOT_ATH_REVENUE/mil)-(TOT_ATH_EXPENSE/mil)), fill=CONFERENCE)) +
  geom_bar(stat='identity', position='identity', color='black', alpha=0.6) +
  labs(title="Athletic Expense vs. Admission Selectivity: Power 5", x="Admission Accept %", y="Athletic Rev - Expense (in Millions)") +
  scale_y_continuous(labels = scales::dollar) +
  theme(axis.text.x=element_text(size=12), axis.title.x=element_text(size=14),
        axis.text.y=element_text(size=12), axis.title.y=element_text(size=14),
        plot.title=element_text(size=20, face="bold"), legend.position = "bottom")
```  

----

```{r other_conf_adm_pct, echo=FALSE, warning=FALSE}

# Difference in Athletic Rev/Exp vs. Selectivity, all other conf.

other_conf %>%
  ggplot(aes(x=Admit_Rate, y=((TOT_ATH_REVENUE/mil)-(TOT_ATH_EXPENSE/mil)), fill=CONFERENCE)) +
  geom_bar(stat='identity', position='identity', color='black', alpha=0.6) +
  labs(title="Athletic Expense vs. Admission Selectivity: All Other Conf.", x="Admission Accept %", y="Athletic Rev - Expense (in Millions)") +
  scale_y_continuous(labels = scales::dollar) +
  theme(axis.text.x=element_text(size=12), axis.title.x=element_text(size=14),
        axis.text.y=element_text(size=12), axis.title.y=element_text(size=14),
        plot.title=element_text(size=20, face="bold"), legend.position = "bottom")
```  

----

```{r top_deficit, echo=FALSE}

# Create calculated column for percent difference in expense and revenue

lgst_deficit <- ncaa %>%
  select(Institution, SUB_DIV, TOT_ATH_REVENUE, TOT_ATH_EXPENSE) %>%
  group_by(Institution, SUB_DIV) %>%
  transmute(REV_EXP_DIFF=((TOT_ATH_REVENUE-TOT_ATH_EXPENSE)/TOT_ATH_REVENUE) * 100)

# top_n or arrange was not functioning
lgst_deficit <- lgst_deficit[order(lgst_deficit$REV_EXP_DIFF),]
lgst_deficit <- lgst_deficit[1:10,]
```  

```{r lgst_def_plot, echo=FALSE}

# Horizontal bar chart, largest % difference in athletic expenses and revenues

lgst_deficit %>%
ggplot(aes(Institution, abs(REV_EXP_DIFF))) + 
  geom_bar(stat='identity', aes(fill=factor(SUB_DIV)), alpha=.7) + 
  coord_flip() + 
  labs(title="Athletic Revenues & Expenses: Largest % Difference", y="-% Difference In Revenue and Expense", fill="Div. I Subdivision") +  
  scale_fill_discrete(breaks=c(1,2,3), labels=c("FBS", "FCS", "w/o FB")) +
  theme(axis.text.x=element_text(size=12), axis.title.x=element_text(size=14),
        axis.text.y=element_text(size=12), axis.title.y=element_text(size=14),
        plot.title=element_text(size=20, face="bold"), legend.position = "bottom")
```

# Conclusion  

- Many different scenarios
- Athletic spending far outpaced by academic expenditures
- Few institutions generate positive revenue
- Most institutions break even on athletic expenses
- More attention should be put onto athletic team subsidies